---
title: "Worksheet 2: Programming Concepts"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

In this worksheet, we will discuss some programming concepts such as conditional statements, functions, or for-loop.

## 1. Conditional Statements

We can use conditional statements for *flow* control. Below are some essential comparison operators for setting conditions in R:

-   `==` means equality

-   `!=` means "not equal"

-   `<` means "less than" while `<=` means "less than or equal to"

-   `>` means "greater than" while `>=` means "greater than or equal to"

-   `x %in% y` is looking for the value of `x` in the object `y`

We can execute some code under certain conditions with the *if-else* structure:

```{r, eval=FALSE}
if (condition) {
  #code to run when condition is TRUE
  } else {
  #code to run when condition is FALSE
}
```

Let's try a few examples:

```{r}
# Set x to be 4
x <- 4  

# Test if x is less than 5 or not
if(x < 5){
  print("This number is less than five!")
} else {
  print("This number is at least five!")
}
```

```{r}
# What if x was 5?
x <- 5  

if(x < 5){
  print("This number is less than five!")
} else {
  print("This number is at least five!")
}
```

```{r}
# Define a character vector
basket <- c('Apple', 'Orange', 'Passion fruit', 'Banana') 

# Is there an apple in the basket?
if("apple" %in% basket){
  print("There is an apple in the basket!")
} else {
  print("There is no apple in the basket!")
}

# Why is that?
# There is Apple, but not apple
```

We can add many conditions!

```{r, eval=FALSE}
if (condition1){
  #code to run when condition1 is TRUE
  code
  } else if (condition2) {
  #code to run when condition2 is TRUE
  code
  }
  } else if (condition3) {
  #code to run when condition3 is TRUE
  code
  }
  } else {
  #code to run when all conditions are FALSE
  code
}
```

#### **Try it! Write a if-else statement that checks: 1) if a number `x` is positive, print "The number is positive", then 2) if a number `x` is negative, print "The number is negative", and finally 3) if x is zero, print "The number is 0".**

```{r}
x = 'abc'
if (is.numeric(x) & x > 0) {
  print("The number is positive")
} else if (is.numeric(x) & x < 0) {
    print("The number is negative")
} else if (is.numeric(x) & x == 0) {
  print("The number is 0")
} else {
    print("The number is 0")
  }
```

A related function that is very useful is `ifelse()`. This function takes three arguments: a conditional statement and two possible answers. If the logical is `TRUE`, the value in the second argument is returned and if `FALSE`, the value in the third argument is returned. It is particularly useful because it works on vectors.

```{r}
# Set x to be a vector containing different integers
x <- c(4,5,0,-5)

# Calculate the inverse of any x: depends if x is equal to 0 or not
ifelse(x != 0, 1/x, NA)
```

Some other useful functions are `any()`, `all()`, and `which()` to check for `TRUE`s in logical values.

```{r}
# Create a logical vector
logics <- c(TRUE, TRUE, FALSE)
```

The `any()` function takes a vector of logical values and returns `TRUE` if any of the entries is `TRUE`.

```{r}
# Are any of the entries true?
any(logics)
```

The `all()` function takes a vector of logical values and returns `TRUE` if all of the entries are `TRUE`. Here is an example:

```{r}
# Are all of the entries true?
all(logics)
```

The `which()` function takes a vector of logical values and returns the positions of the entries that are `TRUE`. Here is an example:

```{r}
# Which entries are true?
which(logics)
```

We can do some operations on logical values! A `TRUE` entry has a value of 1 while a `FALSE` entry has a value of 0. We can count the number of `TRUE` and `FALSE` entries as follows:

```{r}
# Sum of true entries
sum(logics)

# Sum of false entries (opposite of true!)
sum(!logics)
```

#### **Try it! What if we are interested in the proportion of `TRUE` entries? What proportion of cars in mtcars have values of `mpg` greater than 30?**

```{r}
# Write code here
data(mtcars)
mtcars$mpg > 30
cars_logic <- mtcars$mpg > 30
# 3 possible answers
sum(cars_logic) / (sum(cars_logic) + sum(!cars_logic))
sum(cars_logic) / length(mtcars$mpg)
mean(cars_logic)


```

## 2. Functions

As we become more experienced, you will find yourself needing to perform the same operations over and over. Functions are a way to perform repetitive tasks. We can build a function following that structure:

```{r, eval=FALSE}
function_name <- function(inputs){
  # perform operations on inputs and produce output
  code
}
```

Let's define a function for calculating the z-score associated to a value:

$$
z = \frac{X_i - \bar X}{s}
$$

Consider the following data about the heights of a group of students in this class:

```{r}
# The heights are saved as a vector
heights <- c(61,67,64,63,60,66)

# The mean is 
mean(heights)

# and the standard deviation is 
sd(heights)
```

```{r}
# Define a function called zscore to find the zscore of some numeric value
zscore <- function(x,xbar,s){
  # divide the difference between x and the mean by the standard deviation
  (x - xbar) / s
}

# Find your zscore!
zscore(61, mean(heights), sd(heights))
```

#### **Try it! Write a function that computes the reciprocal of any number. Does your function also work for zero?**

```{r}
# Write code here
reciprocal <- function(x) {
  ifelse(x != 0, 1/x, NA)
}
reciprocal(0)

```

## 3. For Loops

Suppose you want to perform the same task over and over. You can use loops to perform iterative tasks with the following structure:

```{r, eval=FALSE}
for(i in sequence){
  # task
}
```

We can do iterations from a sequence or vector of numbers or characters.

First, let's consider all numbers between 1 and 10:

```{r}
# i takes the value of each element in the sequence 1,2,3,...,10
for(i in 1:10){
  print(i)
}
```

What if we would like to square these numbers?

```{r}
# i takes the value of each element in the sequence 1,2,3,...,10
for(i in 1:10){
  print(i^2)
}
```

Printing is nice but what if we wanted to save those values into a vector? We would need to initialize such a vector and define each element in the loop:

```{r}
# Initialize an empty vector
sq1_10 <- c()

# Remember i takes the value of each element in the sequence 1,2,3,...,10
for(i in 1:10){
  # so we can use the values of i to index the new vector
  sq1_10[i] <- i^2 # the ith element of sq1_10 is i^2
}

# Take a look!
sq1_10
```

We can also define a specific vector and go through each element of the vector:

```{r}
# Recall the vector basket
basket <- c('Apple', 'Orange', 'Passion fruit', 'Banana')

# fruit takes the value of each element in basket
for(fruit in basket){
  print(fruit)
}
```

Let's see the zscores of all heights:

```{r}
# Recall the vector heights
heights <- c(61,67,64,63,60,66)

for(h in heights){
  print(zscore(h, mean(heights), sd(heights)))
}
```

If we want to save the values of the different zscores into a vector, we would need to specify how each element will be created:

```{r}
# Initialize the vector zscores
zscores <- c()

# seq_along() creates a sequence of positions! 1,2,3,...,6
seq_along(heights)

# therefore i takes values 1,2,3,...,6
for(i in seq_along(heights)){
  # assign the value of an element in zscores by indexing
  zscores[i] <- zscore(heights[i], mean(heights), sd(heights)) 
}

# Take a look at the zscores vector!
zscores

# What is the sum of these values?
sum(zscores)
```

#### **Try it! Use a for loop to calculate the reciprocal of all numbers between 1 and 10.**

```{r}
# Write code here
for(i in 1:10){
  print(reciprocal(i))
}

reciprocals <- c()
for(i in 1:10){
  reciprocals[i] <- reciprocal(i)
}
reciprocals
```

## 4. Vectorization

For-loops could slow down the computation. We prefer *vectorization* over for-loops since it results in shorter and clearer code. A vectorized function is a function that will apply the same operation on each value of the vectors. Let's find the zscore of each height, using the function on the vector instead:

```{r}
# Apply the function to a vector
zscore(heights, mean(heights), sd(heights))
```

## 5. Don't reinvent the wheel

There are many built-in functions in R that can make your programming easier. Look up the documentation or search forums to see if a function already exists before you build it.

#### **Your turn!**

Consider the dataset `quakes` which contains information about earthquakes occurring near Fiji since 1964. The first few observations are listed below:

```{r}
# Take a look at the data frame
head(quakes)
```

1.  Suppose you want to find out if there was an earthquake with a magnitude greater than 6 in this data frame. What built-in R function could help you quickly look that up?

```{r}
# Write code here
any(quakes$mag > 6)
```

2.  Create a new vector called `mag_cat` to categorize the magnitude of the earthquake as *moderate* if less than 6 and *severe* otherwise.

```{r}
mag_cat <- c()
quakes$mag
seq_along(quakes$mag)

for (i in seq_along(quakes$mag)) {
  if (quakes$mag[i] < 6.0) {
    mag_cat[i] = "moderate"
  } else {
    mag_cat[i] = "severe"
  }
}
  
mag_cat

# same code but in one line
mag_cat <- ifelse(quakes$mag < 6, "moderate", "severe")

```

3.  Using a for-loop, count how many cases of severe earthquakes were reported in the vector you created for the previous question.

```{r}
count_severe <- 0

# count with a loop
for(i in seq_along(mag_cat)){
  if(mag_cat[i] == "severe"){
  count_severe = count_severe + 1
  } else {
  count_severe = count_severe
  }
}

count_severe
```

4.  Define a new function `zscore2` that takes a vector as an input (instead of a value, mean, and sd as we did for `zscore`). Test your function on `quakes$lat`.

```{r}
# Write code here
zscore2 <- function(x){
  # divide the difference between x and the mean by the standard deviation
  (x - mean(x)) / sd(x)
}

# Find your zscore!
zscore2(quakes$lat)
```

5.  Using a for-loop, apply the function `zscore2` to each column in `quakes`, but the last one. *Would it make sense to find the zscore of `stations`?*

```{r}
# Write code here
jy23294 <- quakes 

# Loop other the number of columns
for(i in 1:(ncol(quakes)-1)){
  jy23294[,i] <- zscore2(quakes[,i])
}
quakes

```
