---
title: "Worksheet 3: Visualization Principles"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

## 1. Set up

The easiest way to get `ggplot2` is to install the whole `tidyverse` package which contains many packages that we will use this semester.

```{r}
# If you are using RStudio ON YOUR OWN COMPUTER AND HAVE NEVER USED ggplot2
# Copy and run the line below in your console before running this code chunk
#install.packages("tidyverse")

library(ggplot2)
```

The dataset used in the following code chunk is the built-in dataset `Loblolly`, that contains information about the growth of Loblolly pine trees.

```{r}
# Save the built-in as a data frame in your environment
my_data <- as.data.frame(Loblolly) 

# Show the first 10 rows of the dataset with head()
head(my_data, 10)
```

To get more information about the dataset, run `?Loblolly` in the console. What does each variable represent?

## 2. Creating a `ggplot`

Let's explore the relationship between age and height of Loblolly pine trees.

### a. Defining a plot

The `ggplot()` function will help us build a graph. It specifies the data frame containing the data to explore and the mapping of the *variables* to visual properties on the graph (using `aes()`, aesthetics).

```{r}
# Create a plot to represent height depending on age
ggplot(data = my_data, aes(x = age, y = height))
```

### b. Adding objects

Geoms are the geometric objects (points, lines, bars, etc.) that can be placed on a graph. They are added using functions that start with `geom_`.

```{r}
# Add a geometric object: points
ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point()

# Note the importance of placing "+" correctly
```

It is a good practice to put the new layers on a new line (so be sure to end a line with +).

```{r}
ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point(size = 3) # Make the points bigger with the option size in the geom object
```

Other options such as `color =` and `alpha =` can be applied within a geometric object.

#### **Try it! Use `color =` to make the points blue. Input a value of alpha between 0 and 1. What does alpha control?**

```{r}
ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point(size = 3, color = "red", alpha = 0.3)

# alpha controls opacity (between 0-1). 0 is most transparent
# lets you see which points are more concentrated
```

You can add many layers to the ggplot.

```{r}
ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point(size = 3) + 
  geom_smooth() # Connect data points 
```

#### **Try it! Use the object `geom_line()` instead of smooth. How is that object different?**

```{r}
ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point(size = 3) + 
  geom_line()
```

### c. Grouping

We can add color and different types of object styles under some conditions. Let's put some color by type of seed. We can modify the aesthetics!

```{r}
# Map the variable Seed to color 
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + # color by Seed
  geom_point(size = 3) 
```

We can also differentiate the different seeds by connecting the points to follow a seed's height as its age increases:

```{r}
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) + 
  geom_line(aes(linetype = Seed)) # Change the line type for each seed (aesthetics)
```

Notice that the legend lists seeds in a somewhat random orders. What if we wanted to specify the order in which seeds are colored? Let's double check the class of this variable and redefine it:

```{r}
class(my_data$Seed)
```

Factor is a type of categorical variable that has different level. When grouping, we often prefer to have a categorical variable defined as a factor (if a variable is only defined as character, we can coerce the type with `as.factor()`).

```{r}
# Specify the order of seed for the legend
my_data$Seed <- factor(my_data$Seed, 
                        levels = c("301", "303", "305", "307", "309", 
                                   "311", "315", "319", "321", "323", 
                                   "325", "327", "329", "331"))

# Rebuild the last graph
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) + 
  geom_line(aes(linetype = Seed))
```

#### **Try it! Instead of coloring the points by Seed, color the points by age. What do you notice about the legend?**

```{r}
ggplot(data = my_data, aes(x = age, y = height, color = age)) + 
  geom_point(size = 3) 

# legend is continuous bc age is continuous
# age is class: numeric. you can make it categorical using "as.factor" and make it into a factor class like we did with Seed

```

### d. Controlling scales

Scale functions (which start with `scale_`) allow you to modify the mapping of the variables to some visual characteristics of the plot. Let's change the *x* and *y* axis scaling, and the colors employed.

```{r}
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) + 
  scale_x_continuous(breaks = seq(0, 30, 1)) + # show tick marks for each year
  scale_y_continuous(breaks = seq(0, 65, 5)) + # show tick marks for each 5 ft
  scale_color_brewer(palette = "Blues")        # change the color palette
```

What is wrong with the last scale?
** Maximum # of color types in "Blues" is 9. But we have more than 9 categories, so those are ommitted **

### e. Adding labels

Graphs should be easy to interpret and informative labels are a key element in achieving this goal. The `labs()` function provides customized labels for the axes and legends.

```{r}
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) +
  labs(title = "Relationship between age and height",
       subtitle = "Loblolly pine trees",
       caption = "Kung, F. H. (1986)",
       x = "Age in years",
       y = "Height in Feet",
       color = "Seed of the tree")
```

### f. Faceting

Facets reproduce a graph for each level of another variable (or combination of variables) and are created using functions that start with `facet_`.

```{r}
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) +
  facet_wrap(~Seed) # create a plot for the relationship between age and height for each seed
```

### g. Choosing a theme

We can customize many aspects of our graphs by hand (colors, scales, background color, grid, ...) or we can use some themes other than the default `ggplot` theme.

```{r}
# Use the default theme_minimal()
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) + 
  geom_line() + 
  theme_minimal() # use a minimal theme
```

We can use the function `theme()` to customize a default theme:

```{r}
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) + 
  geom_line() +
  theme(legend.position = "none") # Get rid of the legend
```

Visit this website to find a list of all themes available: <https://ggplot2.tidyverse.org/reference/ggtheme.html>

## 3. Other comments

### a. Mapping

Mappings that are done within the `ggplot()` function apply to all objects. We can also define the mapping within a `geom`. In that case, they only apply only to that specific object.

```{r}
# Placing color mapping in the ggplot function
ggplot(data = my_data, aes(x = age, y = height, color = Seed)) + 
  geom_point(size = 3) + 
  geom_line()
```

```{r}
# Placing color mapping in the geom_point function
ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point(aes(color = Seed), size = 3) + 
  geom_line()
```

What is similar, what is different between these two graphs?

### b. Graphs as objects

Graphs can be saved as an R object and manipulated further:

```{r}
# Save a version of our graph
myplot <- ggplot(data = my_data, aes(x = age, y = height)) + 
  geom_point(aes(color = Seed), size = 3)

# Display the graph
myplot
```

```{r}
# Add some layers to our saved graph
myplot + geom_line()
```

------------------------------------------------------------------------

## **Your turn!**

Recall the built-in dataset `mtcars`.

```{r}
# Save the built-in as a data frame in your environment
my_data2 <- as.data.frame(mtcars)
```

1.  Investigate the dataset using `head()`, `str()`, `?mtcars`, etc. What do the variables represent? Identify their type.

```{r}
# Your turn!
head(my_data2)
str(my_data2)
```

2.  Create a plot:

-   to investigate the effect of the weight of a car `wt` and its type of transmission `am` on the fuel efficiency `mpg`
-   including labels and titles
-   using a theme to create your graph (see a list of available themes here: <https://ggplot2.tidyverse.org/reference/ggtheme.html>)

```{r}
# Your turn!
my_data2$am <- factor(my_data2$am,
                         levels = c("0", "1"))

ggplot(data = my_data2, mapping = aes(x = wt, y = mpg, color = am)) +
  geom_point(size = 3) +
  labs(title = "Effect of weight and transmission on fuel efficiency",
       x = "Weight of car",
       y = "Type of transmission",
       color = "Fuel efficiency (mpg)")

```

3.  Post your plot on the following slideshow with your code in the speakers note on bottom: <https://docs.google.com/presentation/d/10WL2Y3-Ua2siz-b2dgKKS-MTW_GUu6EQtj7GhBPd_5o/edit?usp=sharing>
