---
title: "Worksheet 4: Describing 1 variable"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

## 1. Set up

Remember to call the package `ggplot2` which contains the functions we will use for visualization:

```{r, echo=FALSE}
# Load a package
library(ggplot2)
```

The dataset we will use for this worksheet is `msleep` which is part of the `ggplot2` package. It contains information about mammals, their sleeping habits, and other characteristics. Let's take a look:

```{r}
# Take a look at the dataset
head(msleep)

# More information about the observations, variables, ...
str(msleep)

# Let's save msleep as a data frame in our environment
msleep <- as.data.frame(msleep)
```

## 2. Univariate graphs for categorical variables

We can represent a single categorical variable with a bar graph.

### a. Using a bar graph

Let's investigate if mammals are mainly carnivore, herbivore, insectivore, or omnivore with a bar graph. What we would like to represent on the y-axis is the frequency of each category.

```{r}
# Define the graph with ggplot and basic aesthetics
ggplot(data = msleep, aes(x = vore)) +
  geom_bar() # Use geom_bar()
```

Note: Another way to produce this graph is to use a `stat_` function instead and defining how the "stat" will be represented:

```{r}
ggplot(msleep, aes(x = vore)) +
  stat_count(geom = "bar") # specify the type of object within the stat_ function
```

We can modify some basic aspects of the first graph to make it more readable:

```{r}
ggplot(data = msleep, aes(x = vore)) +
  geom_bar(fill = "blue", color = "black") + # what does fill do compared to color?
  labs(title = "Diet of mammals",
       x = "Diet", y = "Frequency") # add labels
```

Other statistics we can use to describe a categorical variable are the percentages of each category. The height of the bars can represent percents rather than counts:

```{r}
ggplot(msleep, aes(x = vore)) +
  geom_bar(aes(y = after_stat(count/sum(count)))) + # the y-axis represents a ratio
  scale_y_continuous(labels = scales::percent) + # use a scale to display percentage
  labs(title = "Diet of mammals",
       x = "Diet", y = "Percent") # adjust labels
```

### b. Frequency and relative frequency

When reporting statistics about a categorical variable, we often include frequencies / counts and relative frequencies / proportions / percentages. We can use the function `table()` to find those:

```{r}
# Find frequencies
freq <- table(msleep$vore)
freq 

# Find proportions (from frequency table)
prop.table(freq)
```

### c. Handling missing values

As showed in the first few plots, there is a category called `NA`, meaning that there is no information about the diet of the mammal.

```{r}
# Create a new dataset which excludes rows where vore is NA
msleep_nm <- msleep[!is.na(msleep$vore), ] 
msleep_nm
# only select rows with non-missing values: ! and is.na() function 
```

#### **Try it! Ignoring missing values for the diet, visualize the distribution of the different diets with percents, using labels, and adding a different color per diet:**

```{r}
# Write code here
ggplot(data = msleep_nm, aes(x = vore, fill = vore)) +
  geom_bar(aes(y = after_stat(count/sum(count)))) + # the y-axis represents a ratio
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Diet Distribution", x = "Diet", y = "%")
```

### d. Creating a new categorical variable

To differentiate between small and large mammals based on their body weight, let's create a new variable `size`. Let's use the median to separate the lower half of mammals (small) based on their body weight:

```{r}
# Create a new variable with ifelse(condition, if_true, if_false)
msleep_nm$size <- ifelse(msleep_nm$bodywt < median(msleep_nm$bodywt), "Small", "Large")

head(msleep_nm)
```

#### **Try it! Represent the distribution of size. Why does it make sense?**

```{r}
# Write code here
ggplot(data = msleep_nm, aes(x = size)) +
  geom_bar() +
  labs(title = "Diet Distribution", x = "Diet", y = "Count")
sum(msleep_nm$size == "Small")
```

\*\*They both have the same number of elements. 38 small and 38 large"

### e. Dealing with many categories

If there are many categories, the bar graph can become messy because of the overlapping labels:

```{r}
# Many categories for order
ggplot(msleep, aes(x = order, fill = order)) +
  geom_bar()
```

There are at least 2 solutions to fix this!

```{r}
# Solution 1 : Angle the axis labels within the theme
ggplot(msleep, aes(x = order, fill = order)) +
  geom_bar() + 
  theme(
    # Angle the x-axis label to 45 degrees 
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none") # remove the legend
```

```{r}
# Solution 2 : Flip the orientation of the bars
ggplot(msleep, aes(y = order, fill = order)) + # switch x and y
  geom_bar() + 
  theme(legend.position = "none") # remove the legend
```

Note: There is not a direct way to order the bars in increasing/decreasing order in `ggplot` BUT we will learn some more tools in `tidyverse` that will allow us to do that.

When dealing with many categories, it might be a good idea to regroup some categories together. For example, take a look at the conservation variable:

```{r}
# How many categories for conservation?
table(msleep$conservation)
```

The [Wikipedia page about the conservation status of species](https://en.wikipedia.org/wiki/Conservation_status) regroups `en` and `vu` as `Threatened` and `nt`, `cd`, and `lc` as `Lower Risk` (there is no mention of `domesticated` but we can also consider this category as a lower risk).

#### **Try it! Using the function `ifelse()`, create a new variable called `recoded_conservation` which takes values `Threatened` or `Lower Risk`. Make sure to address any issues with the missing values. Visualize the distribution of that new variable.**

```{r}
# Write code here
msleep$recoded_conservation <- ifelse(!is.na(msleep$conservation), 
                                      ifelse(msleep$conservation %in% c("en", "vu"), "Threatened", "Lower Risk"), 
                                      NA)

#another way: get rid of NA values
msleep_nmc <- msleep[!is.na(msleep$conservation), ]
msleep_nmc$recoded_conservation <- ifelse(msleep_nmc$conservation %in% c("en", "vu"), "Threatened", "Lower Risk")

ggplot(data = msleep_nmc, aes(x = msleep_nmc$recoded_conservation)) +
  geom_bar(aes(y = after_stat(count/sum(count)))) + # the y-axis represents a ratio
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Conservation", y = "Percent")
```

## 3. Univariate graphs for numeric variables

We will discuss different representations for a single numeric variable: histograms, density plots, and boxplots.

### a. Histogram

Unlike categorical variables, it doesn't necessarily make sense to visualize what *exact value* is the most common but rather focus on what *range of values* is the most common. We can do that with a histogram:

```{r}
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram()
```

The bins in a histogram define these *ranges of values*. By default, the number of bins is 30. We can adjust the bins by setting the number of bins or by setting the bin width:

```{r}
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(bins = 20) # adjust the number of bins
```

```{r}
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(binwidth = 5)  # adjust the bin width
```

Notice in that last graph that the first bin seems to include negative numbers... That does not make sense! Mammals cannot sleep a negative amount of hours per day. We can adjust how the bins are cut by specifying the center of the first bin:

```{r}
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(binwidth = 5, center = 2.5) + # adjust the center of the first bin
  scale_x_continuous(breaks = seq(0,25,5), # adjust the tick marks of the x-axis
                     limits = c(0,25)) # adjust the min/max tick marks of the x-axis
```

Noticed how the shape of the histogram changes depending on the bins? We usually recommend to have at least 10 different bins to be able to "see" the variation in our data.

#### **Try it! Let's create a better representation of the distribution of `sleep_total`! Adjust the previous histogram: 1) Define bins of 2 hours, starting at 0 and display those values with `scale_x_continuous()`, 2) Show percents instead of counts on the y-axis, 3) Include colors with bars being orange and the contour is blue, 4) Add labels.**

```{r}
# Find max to determine x scale max
max(msleep$sleep_total)

ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(binwidth = 2, center = 1, fill = "blue", color = "orange", 
                 aes(y = after_stat(count/sum(count)))) + # adjust the center of the first bin
  scale_x_continuous(breaks = seq(0,20,2), # adjust the tick marks of the x-axis
                     limits = c(0,20)) + 
  scale_y_continuous(labels = scales::percent)

```

### b. Density plot

An alternative to a histogram is a density plot which technically is a nonparametric method for estimating the probability density function of a continuous random variable (???). But basically, we are trying to draw a smoothed histogram, where the area under the curve equals one.

```{r}
ggplot(msleep, aes(x = sleep_total)) +
  geom_density(fill = "red")  # use the density geom
```

### c. Boxplot

Another graph for numeric variables is a boxplot which represents the 5-number summary of a variable: minimum, first quartile, median, third quartile, maximum.

```{r}
ggplot(msleep, aes(y = sleep_total)) + # we switched to the y-var
  geom_boxplot() # use geom boxplot
```

Note that the x-axis has no meaning. We can remove it:

```{r}
ggplot(msleep, aes(y = sleep_total)) +
  geom_boxplot() + 
  scale_x_discrete( ) # mute the x scale
```

What about the scale on the y-axis? We can adjust the range with `ylim` so that it includes 0:

```{r}
ggplot(msleep, aes(y = sleep_total)) +
  geom_boxplot() + 
  scale_x_discrete( ) + # mute the x scale
  ylim(0,20)
```

#### **Try it! Create a boxplot to represent `bodywt`. Do you notice anything?**

```{r}
# Write code here
ggplot(data = msleep, aes(y = bodywt)) +
  geom_boxplot()
```

### d. Shape, center, and spread

Shape, center, and spread of numeric data are useful concepts in understanding and explaining data. We generally talk about two kinds of shapes: (approximately) symmetric or skewed (with one tail longer than another). The center is used to identify a single value that is representative of a typical value in the data (usually calculating either the mean or median). The spread, also known as the variation, describes how close the data values are to each other (usually calculating either the standard deviation or IQR).

For example, let's compare the distributions of `sleep_total` and `bodywt` by calculating summary statistics and visualizing the histograms:

```{r}
# Visualize sleep_total
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(binwidth = 2, center = 1, 
                 fill = "orange", color = "blue") +
  scale_x_continuous(breaks = seq(0,24,2), limits = c(0,24)) + 
  labs(title = "Distribution of total amount of sleep for 83 mammals")

# Summary statistics of sleep_total
summary(msleep$sleep_total)
sd(msleep$sleep_total, na.rm = TRUE) # standard deviation is not included in summary()
IQR(msleep$sleep_total, na.rm = TRUE) # IQR is not included in summary(). Note: IQR = Q3 - Q1
```

#### **Try it! Create a histogram to represent `sleep_rem` adjusting the bins. Also calculate summary statistics. How do the histograms of `sleep_total` and `sleep_rem` compare? How do the summary statistics compare?**

```{r}
# Write code here
ggplot(msleep, aes(x = sleep_rem)) +
  geom_histogram(binwidth = 1, center = 0.5, 
                 fill = "orange", color = "blue") +
  scale_x_continuous(breaks = seq(0,8,1), limits = c(0,8)) + 
  labs(title = "Distribution of total amount of sleep for 83 mammals")

summary(msleep$sleep_rem)
sd(msleep$sleep_rem, na.rm = TRUE)
IQR(msleep$sleep_rem, na.rm = TRUE)
```

Let's take a look at the distribution of another variable: `bodywt`.

```{r}
# Visualize bodywt
ggplot(msleep, aes(x = bodywt)) +
  geom_histogram(binwidth = 100, center = 50, 
                 fill = "blue", color = "orange") +
  scale_x_continuous(breaks = seq(0,7000,500)) + 
  labs(title = "Distribution of body weight (kg) for 83 mammals")

# Summary statistics of sleep_total
summary(msleep$bodywt)
sd(msleep$bodywt, na.rm = FALSE)
IQR(msleep$bodywt, na.rm = FALSE)
```

Depending on the shape of the distribution, we prefer to report mean+sd *or* median+IQR. Why is that? **sd would take into account outliers, but IQR woulnd't because it takes quartile 3 - quartile 1**

Describe each distribution:

| Variable      | Shape     | Center | Spread |
|---------------|-----------|--------|--------|
| `sleep_total` | symmetric | mean   | sd     |
| `sleep_rem`   | skewed    | median | IQR    |
| `bodywt`      | skewed    | median | IQR    |
