---
title: "Worksheet 6: Joining/Merging"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

## 1. Set up

Let's call the package `tidyverse` which contains the packages `dplyr` which we will use today.

```{r, echo=FALSE}
# Call the tidyverse package
library(tidyverse)

## dplyr package:
# inner_join(): matches pairs of observations
# left_join(): keep all observations in the first dataset
# right_join(): keep all observations in the second dataset
# full_join(): join all data in both datasets
# anti_join(): return all observations missing in second dataset
# semi_join(): return all matching observations from one dataset
# key is the first column that is similar between both datasets
```

Consider the following datasets containing some information about colleges in Texas.

```{r}
# Total enrollment for Fall 2018
college_size <- data.frame(
  college = c("A&M","UT","Texas Tech"),
  enrollment = c(63694, 51684, 37845))

# Location of the college with city
college_city <- data.frame(
  college = c("UT","Texas Tech","U of H"),
  location = c("Austin", "Lubbock", "Houston"))

# Take a look at these datasets
college_size
college_city
```

## 2. Joining datasets

We will discuss different ways to join datasets based on a **key variable** (a variable, or sometimes a set of variables, that defines a unique row in a dataset).

### a. Inner join

Join datasets using `inner_join()` to get the information they have in common, matching the observations with the key variable.

```{r}
# Apply inner_join()
inner_join(college_size, college_city, by = "college")
# by college, so UT and TT are in both sets
```

### b. Left join

Join datasets using `left_join()` to keep information from the "left" dataset and add information from the "right" dataset, matching the observations with the key variable.

```{r}
# Apply left_join()
left_join(college_size, college_city, by = "college")
```

#### **Try it! Add the information of `college_size` to `college_city` with `left_join()`. How does it compare to the joined dataset above?**

```{r}
# Write code here
left_join(college_city, college_size,  by = "college")
```

### c. Right join

This function is doing the opposite of `leftt_join()` so it is not widely used.

```{r}
# Apply right_join()
right_join(college_size, college_city, by = "college")
```

### d. Full join

Join datasets using `full_join()` to keep information from both datasets, matching the observations with the key variable.

```{r}
# Apply full_join()
full_join(college_size, college_city, by = "college")
```

### e. Anti join

We can use `anti_join()` to get information from the "left" dataset for which there is no information in the second, matching the observations with the key variable.

```{r}
# Apply anti_join()
anti_join(college_city, college_size, by = "college")
```

#### **Try it! Find the information of `college_size` that is not in `college_city` with `anti_join()`.**

```{r}
# Write code here
anti_join(college_size, college_city, by = "college")
```

### f. Semi join

What happens if we use `semi_join()`?

```{r}
# Apply semi_join()
semi_join(college_city, college_size, by = "college")

# What is the difference with this code?
semi_join(college_size, college_city, by = "college")
```

## 3. Other options when joining

### a. Auto-suffixing

Consider another dataset reporting the county instead of the city as a location.

```{r}
# Another dataset which reported counties for location
college_county <- data.frame(
  college = c("UT","Texas Tech","U of H"),
  location = c("Travis","Lubbock","Harris"))
college_county
```

What happens if we are joining 2 datasets with the same variable name that is not a key?

```{r}
# Join the two types of locations
left_join(college_city, college_county, by = "college")
```

Any columns that have the same name in both datasets but are not used to join on will be given suffixes .x and .y to make it clear which original dataset they came from (left and right, respectively). You can modify the default suffix:

```{r}
# To give names to the suffix, use `suffix =`
left_join(college_city, college_county, by = "college",
          suffix = c(".city",".county"))
```

### b. No matching key

Some datasets may refer to the same variable with different names. Consider the following dataset reporting universities instead of colleges.

```{r}
# Another dataset that lists universities instead of colleges
college_city2 <- data.frame(
  university = c("UT","Texas Tech","U of H"),
  location = c("Austin", "Lubbock", "Houston"))
college_city2
```

What happens if we are joining 2 datasets that have different names for the key variable?

```{r, eval = FALSE}
# Join by matching "college" with "university"
left_join(college_size, college_city2)
```

We would need to specify the name of the key in each dataset:

```{r}
# Join by matching "college" with "university"
left_join(college_size, college_city2, by = c("college" = "university"))
```

*Note: The order in which we specify the match for the key variable matters (e.g., try to replace the key by `c("university" = "college")`. Only the first name of the key variable is kept in the joined dataset.*

### c. Duplicates

Some datasets may have duplicates for the key variable. Consider the following dataset reporting universities with the same name but in different cities.

```{r}
# Another dataset that contains different locations for UT
college_city3 <- data.frame(
  college = c("UT", "UT", "Texas Tech", "U of H"),
  location = c("Austin", "El Paso", "Lubbock", "Houston"))
college_city3
```

What happens if we are joining datasets that have duplicates for the key variable?

```{r}
# Joining 2 datasets with a duplicate key
left_join(college_size, college_city3, by = "college")

# Is the information contained by this table correct?
```

Looking for duplicates before joining will help us avoid this issue. Let's use the function `distinct()`.

```{r}
# Specify for which variable we may have duplicates
distinct(college_city3, college)

# Add the argument `.keep_all = T` to keep all other variables
distinct(college_city3, college, .keep_all = T)
```

By default, duplicates with higher values (or appearing second in alphabetical order) are removed.

#### **Try it! Consider the following new dataset `college_system` (counting the number of institutions for each college). Keep the information in that new dataset and add all the information contained in `college_city3`. Is the information contained in the resulting dataset correct?**

```{r}
# Another dataset that contains the number of academic institutions in the system
college_system <- data.frame(
  college = c("UT", "Texas A&M", "Texas Tech"),
  nb_institution = c(8, 7, 5))
college_system
```

```{r}
# Write code here
left_join(college_system, college_city3, by = "college")
```

It is sometimes useful to add repeating information for some rows that share the same key. We just need to be careful that it makes sense!

------------------------------------------------------------------------

#### **Your turn!**

A common example for joining different datasets is using the database about flights from New York City. We will focus on 3 datasets from that database.

```{r, echo=FALSE}
# Paste and run the following into your console: install.packages("nycflights13")

# Load package
library(nycflights13)

# Load 3 datasets in your environment
data(flights)
data(airports)
data(planes)
```

1.  Take a quick look at each dataset using `head()`. What does one row represent in each of these datasets?

```{r}
# Your turn!
head(flights)
head(airports)
head(planes)
```

2.  Add the information about the `planes` to the `flights`. What variable should we use as the key?

```{r}
# Your turn!
left_join(flights, planes, by = "tailnum")
```

3.  Let's add the information about the destination `airports` to the `flights`. What variable should we use as the key?

```{r}
# Your turn!
left_join(flights, airports, by = c("dest" = "faa"))
```

4.  Are there any destination `airports` that did not appear in `flights`? Which function should we use to check this?

```{r}
# Your turn!
anti_join(airports, flights, by = c("faa" = "dest"))

```

5.  How many destination airports do appear in `flights`? Which function should we use to check this?

```{r}
# Your turn!
distinct(flights, dest, .keep_all = T)

# there are 105 distinct destination airports
```
