---
title: "Worksheet 8: Data Transformation"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

## 1. Set up

Let's call the package `tidyverse` which contains the packages `dplyr` and `ggplot2` which we will use today.

```{r, echo=FALSE}
# Call the tidyverse package
library(tidyverse)
```

We will use the dataset `txhousing` for this worksheet. It contains information about the housing market in Texas. Run `?txhousing` in your console for more details.

```{r}
# Save dataset as a dataframe in your environment
txhousing <- as.data.frame(txhousing)

# Use glimpse(), a dplyr function similar to str() and head() with the pipe
glimpse(txhousing) 

# filter() chooses only rows you want
```

## 2. The pipe

We will use `dplyr` functions to manipulate our datasets with the pipe `%>%` which can combine a series of operations:

```{r}
# check how the pipe works
1:10 %>% sum() # instead of sum(1:10)

# add as many operations as needed...
1:10 %>% sum() %>% sqrt() # instead of sqrt(sum(1:10))
1:10 %>% sum() %>% sqrt() %>% log() # instead of?
```

*Note: you might see the new pipe in some of the textbooks `|>`. Unfortunately, our server does not have the latest version of R available.*

#### **Try it! Use the pipe to find the maximum year in `txhousing`.**

```{r}
# Write code here
```

## 3. Core `dplyr` functions

### a. Filter

Use `filter()` to choose rows/observations verifying some conditions. Use conditional statements such as `<`, `>`, `>=`, `<=`, `==`, `is.na()`, `&`, `|`, `!`, `%in%`, ...

```{r}
# Filter a categorical variable
txhousing %>%
  filter(city == "Austin")
```

```{r}
# Filter with multiple criteria
txhousing %>% 
  filter(city == "Austin" & sales <= 1000)
```

```{r}
# Filter a numeric variable to a given range
txhousing %>% 
  filter(between(sales,900,1000)) # between(variable, min, max)

# slice() grab rows by row index
txhousing %>% slice(1, n()) # grab first row and last row
txhousing %>% slice(1:3) #grab first three rows
```

#### **Try it! Filter the observations to only display the month of December for the year 2000 and 2010 in Houston.**

```{r}
# Write code here
txhousing %>% filter(city == "Houston" & month == 12 & (year == 2000 | year == 2010))
```

### b. Arrange

Use `arrange()` to sort rows/observations on some conditions. Default is ascending (from least to greatest or alphabetically for categories) but to sort in the other direction use `desc()`.

```{r}
# Sort by median house price, least-to-greatest 
txhousing %>% 
  arrange(median)
```

```{r}
# Sort by city reverse-alphabetically, year and month reverse-chronologically
txhousing %>% 
  arrange(desc(city), desc(year), desc(month))
```

#### **Try it! Sort the observations from least to greatest listings in Dallas for the year of 2012.**

```{r}
# Write code here
txhousing %>% 
  filter(city == "Dallas" & year == 2012) %>%
  arrange(listings)
```

### c. Select

Use `select()` to keep or rename a subset of columns/variables.

```{r}
# Select to keep only some variables
txhousing %>% 
  dplyr::select(city, year, inventory) 
# Note: need to specify the "select" function from the dplyr package if this error occurs:
# Error in select(.,...) unused arguments
```

```{r}
# Select to keep columns using indexes
txhousing %>% 
  dplyr::select(1,2,8)
```

```{r}
# Drop variables using "-"
txhousing %>%
   dplyr::select(-city, -year, -inventory)
```

```{r}
# Move median and inventory to the front
# Use everything() to keep everything else in the same order
txhousing %>% 
  dplyr::select(median, inventory, everything())
```

```{r}
# Select and rename...
txhousing %>% 
  select(Location = city, 
         "Calendar_Year" = year, 
         "Total_Listings" = listings)
```

```{r}
# or just use rename()
txhousing %>% 
  rename(Location = city, 
         "Calendar Year" = year, 
         "Total Listings" = listings)
```

```{r}
# Compare select() and pull()
# select() selects column(s) in data frame
sales_1 <- txhousing %>% select(sales) # results in ?
# pull() pulls the data from the column and makes it a numeric array
sales_2 <- txhousing %>% pull(sales)  # results in ?
class(sales_1) 
class(sales_2)
```

#### **Try it! Let's keep only the variables of `listings` and `sales` for Austin in 2014.**

```{r}
# Write code here
```

### d. Chains

With a pipe `%>%`, we can apply several `dplyr` functions but also other functions:

```{r}
# Chain functions with a pipe after each
txhousing %>% 
  filter(year == 2015, month == 7) %>% # Filter by year 2015, month of July
  dplyr::select(city, median) %>% # Keep city and median
  arrange(desc(median)) %>% # Arrange by descending median
  head(10) # Show first 10 rows
```

After manipulating our dataset, we can build a `ggplot`:

```{r}
# Let's add a ggplot after using the dplyr functions above!
txhousing %>% 
  filter(year == 2015, month == 7) %>% 
  dplyr::select(city, median) %>% 
  arrange(desc(median)) %>%
  head(10) %>%
  ggplot(aes(y = city, x = median)) + 
  geom_bar(stat = "identity")
```

Well, the bars are still not ordered by the value of the median... Let's use a new `dplyr` function within the `gpplot` to order the cities from greatest to least values of the median:  

```{r}
# Reorder the cities by median sale price
txhousing %>% 
  filter(year == 2015, month == 7) %>% 
  dplyr::select(city, median) %>% 
  arrange(desc(median)) %>%
  head(10) %>%
  ggplot(aes(y = reorder(city, median), x = median)) + # reorder the y-axis # reorder(variable, redorder by)
  geom_bar(stat="identity")
```

### e. Mutate

Use `mutate()` to create new columns/variables:

```{r}
# Find the average sale price per row and create new column in dataset
txhousing %>% 
  mutate(average = volume/sales)

# recode numeric variable into character
txhousing %>% mutate(month2 = recode(month, "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) # if month is 1, then it becomes Jan, if month is 2 it becomes Feb
```

#### **Try it! Create a new variable called `position` that is the difference between the `sales` and the average number of `sales`.**

```{r}
# Write code here
txhousing %>% 
  mutate(average = volume/sales) %>%
  mutate(position = average - sales)
```

### f. Summarize

Use `summarize()` (or `summarise()`) to calculate summary statistics on columns/variables. Some useful summary functions: `mean()`, `sd()`, `median()`, `IQR()`, `min()`, `max()`, `n()`, `n_distinct()`, `first()`, `last()`, `nth()`, ...

```{r}
# Find the mean volume
txhousing %>% 
  summarize(mean_vol = mean(volume, na.rm = T))
```

```{r}
# Add more summaries:
# the mean volume, number of rows, and number of distinct cities 
txhousing %>% 
  summarize(mean_vol = mean(volume, na.rm = T), # ignore NA values
            n_rows = n(), # number of rows
            n_cities = n_distinct(city)) # number of distinct values
```

#### **Try it! Find the mean `volume` and mean number of `listings` for Austin in 2015.**

```{r}
# Write code here
txhousing %>% 
  filter(year == 2015 & city == "Austin") %>%
  summarize(mean_volume = mean(volume, na.rm = T), mean_listings = mean(listings))
```

### g. Group by

This is one very important function! It enables us to create subsets and apply a function to all these subsets. For example, find summaries per city and per year:

```{r}
# Find summaries by subgroups
txhousing %>% 
  group_by(city) %>% # subsets per city, per year # separates data into chunks by city
  summarize(mean_vol = mean(volume, na.rm = T), # mean volume
            sd_vol = sd(volume, na.rm = T), # standard deviation
            n = n(), # count observations per subgroup
            se_vol = sd_vol/sqrt(n)) # use recently defined variables

txhousing %>% 
  group_by(city,year) %>% # subsets per city, per year # separates data into by city and year
  summarize(mean_vol = mean(volume, na.rm = T), # mean volume
            sd_vol = sd(volume, na.rm = T), # standard deviation
            n = n(), # count observations per subgroup
            se_vol = sd_vol/sqrt(n)) # use recently defined variables
```

```{r}
# Stop grouping with ungroup()
txhousing %>% 
  group_by(year) %>% # subsets per year
  summarize(count = n()) %>% # count how many observations per year
  ungroup() %>% # ungroup to consider all years
  summarize(n_distinct(count)) # how many of these counts are distinct?
```

------------------------------------------------------------------------

## **Your turn!**

Let's practice some `dplyr` functions on the dataset `txhousing` (think logically).

1.  Is anyone in your group from Texas? Are the cities where your group members are from in the `txhousing` dataset?

2.  Filter rows that contain the cities found in 1. and for the most recent year (using `year == max(year)`). If no one is from Texas, filter 3 cities of your choice.

```{r}
# Your turn!

```

3.  Find the average number of sales per month for each city.

```{r}
# Your turn!

```

4.  Create a plot to represent the average number of sales per month for each city. Add colors and labels to your plot!

```{r}
# Your turn!

```

5.  Post your plot on the following slideshow with your code in the speakers note on bottom: <https://docs.google.com/presentation/d/1iJOCiFsGMf8F6M5olVvvBbT3dnZL3rGLYzdTYj5vQQE/edit?usp=sharing>
