---
title: "Worksheet 9: More Data Manipulation"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

## 1. Set up

Let's call the package `tidyverse`:

```{r, echo=FALSE}
# Call the tidyverse package
library(tidyverse)
```

We will continue using the dataset `txhousing` for this worksheet.

```{r}
# Save dataset as a dataframe in your environment
txhousing <- as.data.frame(txhousing)

# Use head(n) to take a look at the first n rows
txhousing %>%
  head(3)

# Use dim() to find the dimensions of the dataset
txhousing %>%
  dim() 
```

We will continue exploring `dplyr` functions to manipulate our data.

## 2. More examples of filtering

Let's explore a few other functions that can help us choose some rows/observations:

### a. Slicing

```{r}
# Filter rows by indexing
txhousing %>% 
  slice(1:10) # filter the first 10 rows
```

### b. Top values

```{r}
# Select top rows (max values) for a variable
txhousing %>% 
  top_n(3, sales) %>% # top_n(number of rows, variables)
  arrange(desc(sales))
```

```{r}
# Select bottom rows (min values) for a variable
txhousing %>% 
  top_n(-3, sales) %>% # add - for descending order
  arrange(desc(sales))
```

```{r}
# Select top percent of rows (max values) for a variable
txhousing %>% 
  top_frac(0.001, sales) %>% # top_frac(proportion of rows, variables) 
  arrange(desc(sales))
```

#### **Try it! Filter the observations to only display the top 5 number of listings for the city of Austin in descending order.**

```{r}
# Write code here
txhousing %>% filter(city == "Austin") %>%
  top_n(5, listings) %>%
  arrange(desc(listings))
```

### c. Sample values

```{r}
# Take a sample of rows with sample_n()
txhousing %>% 
  sample_n(5) # sample_n(number of rows)
```

By default, the sampling occurs without replacement (cannot pick the same value more than once). You can sample with replacement by adding the argument `replace = TRUE`.

```{r}
# Take a sample of rows with sample_frac()
txhousing %>% 
  sample_frac(.001) # sample_frac(proportion of rows)
```

If you run the code above again, do you get the same sample?

```{r}
# Using a seed
set.seed(320)

# We should all get the same sample!
txhousing %>% 
  sample_frac(.001)
```

Setting a seed, determines a unique "random" process to choose a sample so that we would all get the same!

## 3. More examples of creating new variables

Some options in `mutate()` are really convenient to recode old variables into new variables. For example, we can use `ifelse()`:

```{r}
# Create a new variable under a condition with ifelse()
txhousing %>%
  mutate(month_cat = ifelse(month <= 6, # condition
    "first semester", # value if TRUE
    "second semester")) # value if FALSE
```

When we need to differentiate between more cases, we can use `case_when(condition1 ~ new value, condition2 ~ new value, ...)`:

```{r}
# Create a new variable under different conditions with case_when()
txhousing %>%
  mutate(month_cat = case_when(
    month <= 4 ~ "first trimester",
    4 < month & month <= 8 ~ "second trimester",
    month >= 9 ~ "third trimester"))
```

Or we can use `recode(old_var, new values in order from least to greatest)`:

```{r}
# Recode a numeric variable into a categorical variable
txhousing %>% 
  mutate(month_es = recode(month,
                           "enero", "febrero","marzo",
                           "abril","mayo", "junio",
                           "julio","agosto","septiembre",
                           "octubre","noviembre","diciembre"))
```

Some other examples:

```{r}
# Recode a categorical variable into lower case
txhousing %>% 
  mutate(city = tolower(city)) # use toupper() for upper case
```

```{r}
# Recode a categorical variable by changing the string
txhousing %>% 
  mutate(city = str_replace(city, "Fort", "Ft")) %>% # str_replace(variable, old, new)
  filter(grepl("Ft", city)) # filter cities starting with Ft
```

We will learn more about processing strings next week!

Another useful function for analyzing time series:

```{r}
# For a month's sales, subtracts the previous month's sales with lag(sales))
txhousing %>% 
  filter(city == "Austin") %>% # it only makes sense if we focus on one city at a time
  arrange(year, month) %>% # make sure values are ordered per month and year
  mutate(change = sales - lag(sales)) # lag(): previous value
```

Let's visualize this difference with a ggplot:

```{r}
txhousing %>% 
  filter(city == "Austin") %>%
  arrange(year, month) %>% 
  mutate(change = sales - lag(sales)) %>% 
  # Let's add a ggplot!
  ggplot(aes(x = date, y = change, color = month)) + # use the variable date on the x-axis (combination of month/year)
  geom_line() + 
  scale_color_gradient2(low = "blue", mid = "red", midpoint = 6.5) # color the lines depending on the month with summer months in red
```

#### **Try it! Let's find the annual percentage of sales that occur each month (for example, in Abilene, about 5.2% of the sales in 2000 occurred in January). To find the annual percentage of sales per city: 1) group by the appropriate variable(s), 2) create a variable `annual_sales` to find the total number of sales over a year, 3) create a variable `pct_annual_sales`, dividing the monthly `sales` by the `annual_sales`. Use a `ggplot` to visualize the `pct_annual_sales` depending on the `date` (use the same gradient scale as the previous graph to color `month`).**

```{r}
# Write code here
txhousing %>% 
  group_by(city, year) %>%
  mutate(annual_sales = sum(sales)) %>%
  mutate(pct_annual_sales = sales/annual_sales) %>%
  ggplot(aes(x = date, y = pct_annual_sales, color = month)) + 
  geom_line() + 
  scale_y_continuous(labels = scales::percent) +
  scale_color_gradient2(low = "blue", mid = "red", midpoint = 6.5) # color the lines depending on the month with summer months in red

txhousing %>% 
  group_by(city, year) %>%
  mutate(total_price = sum(median)) %>%
  mutate(pct_total_price = median/total_price) %>%
  ggplot(aes(x = date, y = pct_total_price, color = month)) + 
  geom_line() + 
  scale_y_continuous(labels = scales::percent) +
  scale_color_gradient2(low = "blue", mid = "red", midpoint = 6.5) # color the lines depending on the month with summer months in red
  
```

## 4. Suffixes

We can add suffixes to the core functions of `summarize` or `mutate`:

-   `_all()`: applies to every variable\
-   `if()`: applies to variables when a conditional statement is TRUE.\
-   `_at()`: applies to some selected variables (using `c("var1", "var2")` or `vars(var1,var2)`)\

### a. Summarize

Let's `summarize` with these suffixes! Summarizing all variables:

```{r}
# Compute the mean for all variables
txhousing %>% 
  summarize_all(mean, na.rm = T)
```

```{r}
# Find the number of distinct values (ignoring missing data) of each variable
txhousing %>% 
  summarize_all(n_distinct, na.rm = T)
```

However, it doesn't make sense to calculate means for categorical variables and find distinct values for numeric variables. Let's use `summarize_if()` instead!

```{r}
# Summarize variables that meet certain criteria
txhousing %>% 
  summarize_if(is.numeric, mean, na.rm = T) # Only compute the mean for numeric variables
```

```{r}
txhousing %>% 
  summarize_if(is.character, n_distinct, na.rm = T) # Only find the number of distinct values for categorical variables
```

Or we could be interested in just some particular variables:

```{r}
# Only summarize for a few variables: all of the following are equivalent!
txhousing %>% 
  summarize_at(c("sales","volume","median"), mean, na.rm = T) # using c()

txhousing %>% 
  summarize_at(vars(sales:median), mean, na.rm = T) # using vars()

txhousing %>% 
  summarize_at(4:6, mean, na.rm = T) # using indexing
```

### b. Mutate

Let's `mutate` with suffixes!

```{r}
# Coerce all variables into categorical
txhousing %>% 
  mutate_all(as.character)
```

```{r}
# Make all character variables as factors
txhousing %>% 
  mutate_if(is.character, as.factor)
```

```{r}
# Make some numeric variables as factors
txhousing %>% 
  mutate_at(c("year","month"), as.factor)
```

#### **Try it! Create a function to standardize numeric values (complete code below). Then standardize the variables of `sales` to `inventory` in the `txhousing` dataset. Finally, find the sum of the standardized variables.**

```{r}
# Create a function to standardize our values
zscore <- function(x) {
  (x - mean(x, na.rm = T))/ sd(x, na.rm = T)
}

# Write code here
txhousing %>% 
  mutate_at(vars(sales:inventory), zscore) %>%
  mutate_at(vars(sales:inventory), sum)
  
```

## 5. When to `dplyr` or not `dplyr`

Let's compare a graph after using `dplyr`:

```{r}
txhousing %>% 
  group_by(year,month) %>% # create subsets by year, month
  summarize(mean_vol = mean(volume, na.rm=T), # mean volume
            sd_vol = sd(volume, na.rm=T),     # standard deviation  
            n = n(),                          # number of months in a year
            se_vol = sd_vol/sqrt(n)) %>%      # standard error
  ggplot(aes(x = month, y = mean_vol, fill = month)) + 
  geom_bar(stat = "identity") +               # represent the mean volume
  geom_errorbar(aes(y = mean_vol,             # include error bars
                    ymin = mean_vol - se_vol, # define min and max of error bars
                    ymax = mean_vol + se_vol)) +
  facet_wrap(~ year) + 
  scale_x_continuous(breaks = seq(1,12,2)) +  # tick mark for every 2 months
  scale_fill_gradient2(low = "blue", mid = "red", midpoint = 6.5)
```

And just a `ggplot`:

```{r}
ggplot(txhousing, aes(x = month, y = volume, fill = month)) + 
  geom_bar(stat = "summary", fun = mean) +                # represent the mean volume
  geom_errorbar(stat = "summary", fun.data = "mean_se") + # represent error bars with SE
  facet_wrap(~ year) + 
  scale_x_continuous(breaks = seq(1,12,2)) +
  scale_fill_gradient2(low = "blue", mid = "red", midpoint = 6.5)
```

Same graph but less code!

------------------------------------------------------------------------

## **Your turn!**

Let's practice our skills with some joining functions on the dataset `txhousing` to join some weather data!

1.  We will first upload a dataset containing information about the weather in Austin, TX in 2014. How many observations and variables are there in the dataset? How do you identify a unique observation in this data (i.e., what are the key variables)?

```{r}
# Upload the weather dataset from my github to your environment
weather <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//ATX_weather_2014.csv")

# Take a look!
weather %>% 
  head()
```

2.  Since the `txhousing` dataset contains housing data *month by month*, we would also like to have *monthly* weather data. To do so, compute the mean of the `weather` data (temperature, pressure, precipitation), using `group_by()` and `summarize_at()`. Don't forget to add `na.rm = T` in case there are some missing values! Save the new dataset as `atxweather`.

```{r}
# Your turn!
atxweather <- weather %>%
  group_by(year, month) %>%
  summarize_at(vars(temperature:precipitation), mean, na.rm = T)
  

```

3.  Filter the `txhousing` dataset to only retain information about Austin in 2014. Save the new dataset as `atxhousing`.

```{r}
# Your turn!
atxhousing <- txhousing %>% 
  filter(city == "Austin", year == 2014)

```

4.  What variable do `atxweather` and `atxhousing` have in common to identify a unique observation? Join these datasets using any of the joining functions `inner_join`, `left_join`, `right_join`, or `full_join` (all should result in the same data: why?) with the variable in common as a key. Call the resulting dataset `atx`.

```{r}
# Your turn!
atx <- left_join(atxhousing, atxweather, by = c("month", "year"))

```

5.  Using the joined dataset, create a plot to investigate the impact of the weather on the housing market (i.e., pick one `atxweather` variable and one `atxhousing` variable). Add colors and labels to your plot!

```{r}
# Your turn!
atx %>%
  ggplot(aes(x = temperature, y = sales, color = temperature)) +
  geom_point() +
  geom_line() +
  labs(x = "Temperature", y = "Sales", title = "Sales vs. Temp") +
```

6.  Post your plot on the following slideshow with your code in the speakers note on bottom: <https://docs.google.com/presentation/d/1tNKF8tvvB8OeMkqITytOF0P9oh5J9U5IWJGdDjxgKb4/edit?usp=sharing>

*Note: data was retrieved from <https://www.ncei.noaa.gov/products/land-based-station/integrated-surface-database>*
