---
title: "Worksheet 10: More Wrangling"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print = 10))

# Edit the file starting below
```

## 1. Set up

Let's call the package `tidyverse` which contains the packages we will need for today. Let's also use the package `nycflights13` to access the database about flights from New York City. We will focus on 5 datasets in that database.

```{r, echo=FALSE}
# Paste and run the following into your console: install.packages("nycflights13")

# Load packages 
library(tidyverse)
library(nycflights13)

# Call 5 datasets in your environment
data(flights)
data(airports)
data(airlines)
data(planes)
data(weather)
```

## 2. Missing data

In R, the philosophy is that missing values should never silently go missing. That's why if there is any missing value in the input, the output will be a missing value.

For example, we have come across that issue many times by now, what happens if we calculate the mean of a variable with some missing values?

```{r}
# Calculate the mean of departure time
mean(flights$dep_time) # what should we do to get the mean of the non-missing values?
```

Same thing happens when we are making graphs:

```{r}
# How do missing values appear in a histogram?
ggplot(flights, aes(x = dep_time)) +
  geom_histogram(binwidth = 100, center = 50, 
                 fill = "lightblue", color = "blue") +
  scale_x_continuous(breaks = seq(0,2400,200))

# How do missing values appear in a bar graph?
ggplot(airports, aes(y = tzone)) +
  geom_bar(fill = "lightblue")
```

It would be convenient to know which variables have missing values and find the proportion of missing values.

#### **Try it! Complete the following code to find the proportion of missing values for all variables.**

```{r}
flights %>%
  mutate_all(is.na) # continue the code
```

The variable of `dep_time` has about 2.45% of missing values. We could decide to remove missing data for this variable for example. But what might the flights with missing departure time represent?

```{r}
# Take a look at the missing values for departure time
flights %>% 
  filter(is.na(dep_time))
```

What if we would like to remove the rows with missing values?

```{r}
# Remove missing values for the variable of dep_time
flights %>% 
  filter(!is.na(dep_time))
```

We could also choose to remove all rows containing any missing values (for any variable):

```{r}
# All of the following are equivalent!

# using na.omit()
flights %>% 
  na.omit()

# using complete.cases()
flights %>% 
  filter(complete.cases(.))

# using drop_na()
flights %>% 
  drop_na()
```

But be careful about removing all `NA` values as sometimes those have a meaning!

## 3. Outliers

Let's discuss two examples of outliers.

First example appears in the `weather` dataset for `wind_speed` (which is expressed in mph). We best spot outliers with a visualization:

```{r}
# Use a histogram
ggplot(weather, aes(x = wind_speed)) +
  geom_histogram()

# Outliers best appear in a boxplot
ggplot(weather, aes(y = wind_speed)) +
  geom_boxplot() + 
  scale_x_discrete()
```

What do you think about that extreme value of wind speed? How does it affect the calculation of the mean wind speed for example?

#### **Try it! Remove the extreme value of wind speed and visualize the distribution with a boxplot. What do you notice in that plot?**

```{r}
# Write code here
```

Second example of outliers appears in the `planes` dataset for `seats` (number of seats in a plane). As mentioned above, best way to spot outliers is with a visualization:

```{r}
# Spot outliers in a boxplot
ggplot(planes, aes(y = seats)) +
  geom_boxplot(alpha = 0.5) + 
  scale_x_discrete()
```

What do you think about the extreme values of seats? Is it reasonable to drop them? Why/Why not?

*Note: to detect missing data or outliers, we use descriptive statistics and visualizations. This is a crucial step for understanding our data and its structure.*

------------------------------------------------------------------------

## **Your turn!**

You will practice some data wrangling with the NYC flights with minimum guidance!

1.  Show the carriers (display the names) from the shortest to longest mean departure delay. What other descriptive statistic is usually reported with the mean? Include it in the report as well as the minimum value. Interpret the minimum values in context.

```{r}
# Your turn!

```

2.  Look up the flight for the minimum value for departure delay. What is the arrival delay for this flight? It might look suspicious! To check if that could be an error, display the length of the flight (called `air_time`) with a boxplot for that specific route (specify airport of origin and destination). Do you think that there was an error recording that arrival delay?

```{r}
# Your turn!

```

3.  Recall the meaning of missing departure time. Which airport (display the full name) had the biggest number of cancelled flights in 2013?

```{r}
# Your turn!

```

4.  Instead of how frequently flights are cancelled at each airport, we should report the percentage of flights that are cancelled at each airport. How would we go about calculating such a percentage? *This is a little challenging.*

```{r}
# Your turn!

```

5.  Pick one of these datasets: airports, airlines, planes, weather. Explore missing values in these datasets: do the missing values, if any, have a meaning? Is there a reason why they should be missing? Also investigate some of the variables to see if there are some outliers. Report your findings in this slideshow: <https://docs.google.com/presentation/d/1BgOr53HECamMT5Jip7b2Sfzjw1hA-PmuQ5a8GeCtqPo/edit?usp=sharing>

```{r}
# Your turn!

```


