---
title: "Worksheet 11: String Processing"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print = 10))

# Edit the file starting below
```

## 1. Set up

Let's call the package `tidyverse` that also contains the `stringr` package.

```{r, echo=FALSE}
# Load package
library(tidyverse)
```

We will manipulate strings with functions that have intuitive names, and all start with `str_`. We will refer to a built-in character vector called `fruit` for practicing string manipulations:

```{r}
# Example of a built-in character vector
fruit
```

## 2. Strings

### a. Create strings

We can create strings with either single quotes `'` or double quotes `"`.

```{r}
"Here is an example string"

'Here is an another "example" string'
```

To escape a special character such as quotes, we will use a backslash `\`:

```{r}
# Single quote
'\''
```

We can store several strings into a vector with `c()`:

```{r}
c("a", "b", "c")
```

### b. Length of a string

Let's use `str_length()` to find the length of a string:

```{r}
# String length
str_length("Spring Break") # a space counts as a character
```

We can apply this function to many strings!

```{r}
# String length of the built-in vector
fruit %>% str_length
```

### c. Combine strings

We can use `str_c()` to combine two or more strings:

```{r}
# Combine strings
str_c("Spring", "Break")

# By default, no space but we can add the argument sep = 
str_c("Spring", "Break", sep = " ")
```

```{r}
# Add a prefix by using a vector in the second argument
str_c("prefix-", c("a","b","c"))

# Add a suffix by using a vector in the first argument
str_c(c("a","b","c"), "-suffix")

str_c("prefix-", c("a","b","c"), "-suffix")
```

What if we want to combine all the values of one vector together?

```{r}
# Use the argument collapse =
str_c(c("a","b","c"), collapse = "")

# Or separate by a comma and a space
str_c(c("a","b","c"), collapse = ", ")
```

### d. Modify strings

We can change the strings from lower to uppercase and vice-versa (also use sentence case):

```{r}
# To lower case
str_to_lower("Spring Break")

# To upper case
str_to_upper("Spring Break")

# To sentence case
str_to_sentence("Spring Break")
```

### e. Subset strings

We can also focus on a subset of a string with `str_sub()` (only works with positions):

```{r}
# Select a position in the string
fruit %>% str_sub(start = 1, end = 3)

# Or count backwards with -
fruit %>% str_sub(start = -3, end = -1)
```

## 3. Regular expressions (Regex)

Regular expressions are used to describe patterns in strings. They're a little weird at first but they can be very useful!

### a. Basic match for patterns

We can view and detect match patterns with `str_view()`, `str_view_all()`, and `str_detect()`.

```{r}
# View the match(es) of "an"
fruit %>% str_view("an")

# View all possible matches within one word 
fruit %>% str_view_all("an")
```

What does `str_detect` do differently?

```{r}
# Detect the matches
fruit %>% str_detect("an")
```

#### **Try it! Count how many fruits contain the pattern `berry`.**

```{r}
# Write code here
fruit %>% 
  str_detect("berry") %>%
  sum
  
```

### b. Wildcards

Use `.` that matches any character (except a new line):

```{r}
# View all matches with any letter, a, and any letter (period matches with any letter)
str_view_all(fruit,".a.")
```

### c. REGEX Anchors

Let's find a match at the beginning of a string with `^` or at the end of a string with `$` :

```{r}
# Match at the beginning with "a" (^a denotes beginning with a)
fruit %>% str_view("^a")

# Match at the end with "a" (a$ denotes ending with a)
fruit %>% str_view("a$")
```

#### **Try it! What is the proportion of fruit names that start with the first letter of your name?**

```{r}
# Write code here
fruit %>% 
  str_detect("^s") %>%
  mean

```

### d. Literal characters

What if our string include special characters? We need to use an "escape". For example, `\\` means literal `\` in strings.

```{r}
# Test weird strings
weirds <- c("a.A", "a\\c", "a/c", "a c") 

# Match literal period. Need two backslash for period
str_view(weirds, "\\.")

# anything, period, anything
str_view(weirds, ".\\..") # note the distinction with this command

# Match literal backslash
str_view(weirds, "\\\\")
```

### e. Special patterns

Here are a few useful patterns:

-   `[0-9]` matches any digit

-   `[ ]` matches any single space

-   `[^ ]` matches anything except a space

-   `[abc]` matches a, b, or c

-   `[^abc]` matches anything except a, b, or c

-   `[a-zA-Z]` matches any letter

-   `[a-zA-Z0-9]` matches any alphanumeric

```{r}
# What do these lines do? 
str_view_all(weirds,"[ ]")
str_view_all(weirds,"[aeiou]")
str_view_all(weirds, "[^a-zA-Z]")
```

#### **Try it! Show all vowels, lower or upper case in `weirds`.**

```{r}
# Write code here
str_view_all(weirds, "[aeiouAEIOU]")
```

### f. Repetitions

To control how many times a pattern matches:

-   `?` is for 0 or 1 match
-   `+` is for 1 or more matches
-   `*` is for 0 or more matches
-   `{n}` is for exactly n matches
-   `{n,}` is for n or more matches
-   `{,m}` is for at most m matches
-   `{n,m}` is for between n and m matches

```{r}
# Test repeats
repeats <- c("","z","zz","zzz")

# What do these lines do? 
str_view(repeats, "z?")
str_view(repeats, "z+")
str_view(repeats, "z*")
str_view_all(repeats, "z{2}")
```

### g. Capture groups

We can capture groups with regular expressions using `()`:

```{r}
# Remember: 
# . refers to any character 
# + refers to potential repeats (1 or more times)
fruit %>% str_view("(.+)berry") # any letters before berry
```

Let's learn a new `stringr` function called `str_match()`.

```{r}
# What does str_match() do?
fruit %>% str_match("(.+)berry")
```

#### **Try it! After finding the matches of the words ending with `berry`, define the resulting object `as.data.frame`, omit missing values, and only keep the names of the fruits.**

```{r}
# Write code here
fruit %>% 
  str_match("(.+)berry") %>%
  as.data.frame %>%
  na.omit %>%
  select(V1)
```

That was a lot of work! Luckily, `str_subset()` directly subsets matching strings as a vector!

```{r}
# use str_subset() instead
fruit %>% str_subset("(.+)berry")
```

### h. Replace pattern

Let's try `str_replace()` and `str_replace_all()`:

```{r}
# Look for a pattern...
str_view("banana", "an")

# ... and replace that pattern
str_replace("banana", "an", "oon")
```

```{r}
# Look for all patterns...
str_view_all("banana", "an")

# ... and replace these patterns
str_replace_all("banana", "an", "oon")
```

Sometimes missing values were not coded using the R convention of `NA`. Here is an example on how to identify missing values using the convention `NA` with `na_if()`:

```{r}
# Create a dataframe with missing values coded as N/A
data.frame(var = c(1, 2, "N/A", "N/A"))

# Replace N/A by the convention NA
data.frame(var = c(1, 2, "N/A", "N/A")) %>% 
  mutate(var_na = na_if(var, "N/A")) %>%
  mutate(var_num = as.numeric(var_na))
```

------------------------------------------------------------------------

## **Your turn!**

Let's practice some string manipulation with another built-in character vector:

```{r}
# Consider this other built-in character vector
sentences %>% head 
```

Notice some of the words in `sentences` are fruits. Let's find which sentences mention fruits!

1.  We will consider all fruit names from `fruit` and look for matches in the `sentences`. Combine all the elements from `fruit` into a single string called `testfruit` that contains all the possible names separated by `|`. What does `|` mean in R anyway?

```{r}
# Your turn!
testfruit <- str_c(fruit, collapse = "|")

sentences %>%
  str_view(testfruit)
```

2.  Use `str_subset()` to find a subset of `sentences` containing a fruit name. Does there appear to be any issue? Look at the third sentence for example.

```{r}
# Your turn!
sentences %>% str_subset(testfruit)


```

3.  Let's try another method: split the sentences as words then look for exact matches. First, get rid of punctuation in the sentences with `str_remove_all()`. *Hint: punctuation is anything that is not a letter or number.* Then convert all characters as lower case and split `sentences` into individual words with `str_split(" ") %>% unlist`. Save the resulting vector as `sentences_words`. What happens if you don't do `unlist`?

```{r}
# Your turn!
sentences_words <- str_remove_all(sentences, "[^a-zA-Z0-9 ]") %>%
  str_to_lower() %>%
  str_split(" ") %>%
  unlist


```

4.  Use the function `which` to find which fruits appear in `sentence_words`.

```{r}
# Your turn!
which(fruit %in% sentences_words)
fruit[which(fruit %in% sentences_words)]

```
