---
title: "Worksheet 5: Describing Relationships"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

## 1. Set up

Remember to call the package `ggplot2` which contains the functions we will use for visualization:

```{r, echo=FALSE}
# Load a package
library(ggplot2)
```

We will continue working with the dataset `msleep` and `msleep_nm` from the previous worksheet:

```{r}
# Let's save msleep as a data frame in our environment
msleep <- as.data.frame(msleep)

# Create a new dataset which excludes rows where vore is NA
msleep_nm <- msleep[!is.na(msleep$vore), ] 

# Create a new variable with ifelse(condition, if_true, if_false)
msleep_nm$size <- ifelse(msleep_nm$bodywt < median(msleep_nm$bodywt), "Small", "Large")
```

The type of graph we will use depends on the types of the variables we want to represent (categorical or numeric), and how many.

## 2. Comparing two categorical variables

### a. Stacked bar graph

We can create stacked bar graphs that compares the distribution of two categorical variables:

```{r}
# Represent the diet on the x-axis and the size as a color
ggplot(msleep_nm, aes(x = vore, fill = size)) +
  geom_bar()
```

### b. Side-by-side bar graph

Or design a side-by-side bar graph with the option `dodge`:

```{r}
ggplot(msleep_nm, aes(x = vore, fill = size)) +
  geom_bar(position = "dodge") # create a side-by-side bar graph
```

### c. Segmented bar graph

We can display the relative distribution at each level with a segmented bar graph:

```{r}
ggplot(msleep_nm, aes(x = vore, fill = size)) +
  geom_bar(position = "fill") # create a segmented bar graph
```

Note that the default label for the y-axis is count. But what does it represent?

#### **Try it! Adjust the label for the y-axis in the previous segmented bar graph. Also swap the variables for `x =` and `fill =`.**

```{r}
# Write code here
ggplot(msleep_nm, aes(x = size, fill = vore)) +
  geom_bar(position = "fill") + # create a segmented bar graph
  labs(y = "proportion")
```

### d. Frequency and relative frequency

We can report frequencies/relative frequencies by crossing two categorical variables. We can use the function `table()` and `prop.table()` to find those:

```{r}
# Find frequencies
freq <- table(msleep_nm$vore, msleep_nm$size)
freq 

# Find proportions (from frequency table): What is different between those two tables
prop.table(freq, 1) # proportion between each row
prop.table(freq, 2) # proportion between each column
```

## 2. Comparing a categorical variable with a numeric variable

We will discuss many different ways to compare a categorical variable with a numeric variable (and vice-versa).

### a. Special case of bar graph

For example, we can represent the mean value of a numeric variable for each level of a categorical variable using `stat` options:

```{r}
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_bar(stat = "summary", fun = "mean") 
# represent a summary stat using the mean function 
```

To represent how spread out the values are within a category, we can add error bars with `geom_errorbar()` (representing +/- 1 standard deviation from the mean):

```{r}
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_bar(stat = "summary", fun = "mean") +
  geom_errorbar(stat = "summary", fun.data = "mean_se")
```

### b. Comparative boxplots

Using comparative boxplots is another way to visualize the relationship between a categorical variable and a numeric variable.

```{r}
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_boxplot()
```

Recall the density plots introduced for a single numeric variable? Violin plots are similar to density plots, but are mirrored and rotated 90 degrees:

```{r}
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_violin() # use when you have multiple categories

```

#### **Try it! Overlap a boxplot inside the previous violin plot. Use the option `width =` to control the width of the boxplot (input a value between 0 and 1 in increments of 0.1). What maximum width keeps all boxplots inside the violin plot?**

```{r}
# Write code here
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_violin() +
  geom_boxplot(width = 0.1)
```

### c. Comparative histograms

Using comparative histograms is a little harder to read:

```{r}
ggplot(msleep_nm, aes(x = sleep_total, fill = vore)) +
  geom_histogram(binwidth = 2, center = 1) +
  scale_x_continuous(breaks = seq(0,20,2))
```

Pretty ugly! But what if we create one histogram for each diet? We can do that with faceting:

```{r}
ggplot(msleep_nm, aes(x = sleep_total, fill = vore)) +
  geom_histogram(binwidth = 2, center = 1) +
  scale_x_continuous(breaks = seq(0,20,2)) +
  facet_wrap(~vore, ncol = 1) # facet per diet and show all graphs in 1 column
```

### d. Reporting center and spread for each category

For now, we haven't learned the tools for the best way to do this. But we'll learn soon!

## 3. Comparing two numeric variables

### a. Scatterplot

The relationship between two quantitative variables is typically displayed using scatterplots or sometimes line graphs (see WS3). Let's compare `brainwt` to `bodywt`:

```{r}
# Relationship between bodywt and brainwt
ggplot(msleep, aes(x = bodywt, y = brainwt)) +
  geom_point(size = 3)
```

### b. Dealing with a large range of values

We already discussed some features of `geom_point()` and `geom_line()` in the previous worksheet WS3. Let's discuss how to improve this particular scatterplot because there are two mammals that distort the view of the rest of the data points (the weights are much higher than other values):

```{r}
msleep[msleep$bodywt > 2000, ]
```

We can adjust the scale on the x-axis and y-axis to focus on the area where most of the points are:

```{r}
ggplot(msleep, aes(x = bodywt, y = brainwt)) +
  geom_point(size = 3) +
  xlim(0,800) + ylim(0,1.5) # Define new limits for the x- and y-axis
```

But that means we are excluding information about these two mammals. Since there is a high variation between the weights, we can consider a logarithmic scale for both the body weight and brain weight:

```{r}
ggplot(msleep, aes(x = bodywt, y = brainwt)) +
  geom_point(size = 3) +
  scale_x_log10() + scale_y_log10() # consider a log scale for both axes
```

Right now the axes appear as a scientific notation. We can change the threshold for scientific notation:

```{r}
# Changing threshold for scientific notation
options(scipen = 999)

# Redo the graph
ggplot(msleep, aes(x = bodywt, y = brainwt)) +
  geom_point(size = 3) +
  scale_x_log10() + scale_y_log10()
```

Using a logarithmic scale, it is easier to visualize the relationship between the two variables!

### c. Correlation

Correlation describes the strength of a (linear) relationship between two variables. We commonly refer to the Pearson correlation coefficient which takes values between -1 (strong negative relationship) and 1 (strong positive relationship) with 0 indicating that there is no relationship.

```{r}
# Find the correlation between two variables
cor(msleep$bodywt, msleep$brainwt, 
    use = "pairwise.complete.obs") # ignore missing values
```

#### **Try it! Calculate the correlation coefficient between `sleep_total` and `awake`. Why does it make sense?**

```{r}
# Write code here
cor(msleep$sleep_total, msleep$awake, 
    use = "pairwise.complete.obs")
```

## 4. Graphs with more than 2 variables

We can display the relationships between more than 2 variables but it can become tricky. There are many ways to adapt graphs to display more than two variables: grouping by color, size, shape, using faceting, etc.

#### **Try it! Create a scatterplot to investigate the relationship between `sleep_total` and `bodywt` (the latter being expressed in a log scale). Color the points by the different diets and control the size of the points depending on the `sleep_rem` of a mammal. Remember to add labels!**

```{r}
# Write code here
ggplot(data = msleep_nm, mapping = aes(x = bodywt, y = sleep_total, color = vore, size = msleep_nm$sleep_rem)) +
  geom_point() +
  scale_x_log10()
  
```

## 5. Choosing an appropriate graph

When deciding how to represent your data, think about what type of data is available. One great website I found to help you decide between tons of visualizations is <https://www.data-to-viz.com/> (made in France!)

## 6. Color palettes

The default colors created by ggplot can be replaced by calling some color palette. Here is a list of the different ones that exist:

```{r}
RColorBrewer::display.brewer.all(type="seq")
```

A few examples with `scale_fill_brewer()` when we are mapping a variable to the aesthetic `fill =` (similarly, we would use `scale_color_brewer()` when we are mapping a variable with `color =`):

```{r}
# Shades of blue
ggplot(msleep_nm, aes(x = vore, fill = vore)) + 
  geom_bar() +
  scale_fill_brewer(palette = "Blues")

# Shades of orange
ggplot(msleep_nm, aes(x = vore, fill = vore)) + 
  geom_bar() +
  scale_fill_brewer(palette = "Oranges")
```

We can also define our own colors with `scale_..._gradient()` with setting the color for *low* values and another color for *high* values:

```{r}
# Represent the amount of total sleep vs log body weight
ggplot(msleep_nm, aes(x = bodywt, y = sleep_total, 
                      color = log10(bodywt))) +
  geom_point() + 
  scale_x_log10() +
  # choose colors of the scale from yellow to red
  scale_color_gradient(low = "yellow", high = "red")
```

Or we can create our own set of colors!

```{r}
# List of color blind friendly colors
color_blind_friendly <- c("#999999", "#E69F00", "#56B4E9",
                          "#009E73", "#F0E442", "#0072B2",
                          "#D55E00", "#CC79A7")

# Same graph as before but color by diet
ggplot(msleep_nm, aes(x = bodywt, y = sleep_total, 
                      color = vore)) +
  geom_point() + 
  scale_x_log10() +
  # choose colors saved in our object
  scale_color_manual(values = color_blind_friendly)
```
