---
title: 'What Major Should You Choose? Comparison of Median Salary Based on Career Status, Salary Growth, and Unemployment Rates'
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
---

```{r global_options, include=FALSE}
# Document setup
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print = 10))

```

# 1. Introduction

One of the most important life decisions an individual makes occurs when he/she is just a high school student: what major to pursue. What major and subsequent degree one obtains is critical for a job after college, and thus setting their career trajectory. Therefore, data was collected and visualized to see which college degrees offer the most competitive pay, so that students are able to make a more educated, data-driven decision.

First, the distribution of the median starting and mid-career salaries of each major/degree category were visualized. Then, plots of the top 5 majors of each category were created based on the median salary. Secondly, salary growth as one progressed from starting to mid-career salary was recorded. Using this data, percent change in median starting salary to mid-career salary for each of the major categories was visualized. Then, the top 20 majors by salary percent change were plotted. Lastly, unemployment rates for each major category was observed and visualized in comparison to the national unemployment rate.

Data for salary distributions and salary growth were obtained from the FiveThirtyEight data repository (Link: [fivethirtyeight/data](https://github.com/fivethirtyeight/data/tree/master/college-majors)), which contained aggregated data from 2010 - 2012. The variables of interest included: the major, major category, total number of people graduated with the major, sample size, number employed, number unemployed, unemployment rate, median salary, P25 salary, and P75 salary. The dataset contained data of the aforementioned variables for both recent graduates (starting career, ages \< 28) and graduates (mid-career, ages 28+).

In addition, change in median salary over the past decade, and unemployment rate comparison to the national unemployment rate (for 2021) was visualized. In order to extract the required data, the ACS 1-Year PUMS data from the United States Census Bureau (Link: [PUMS data](https://www.census.gov/programs-surveys/acs/microdata/access.2021.html#list-tab-735824205)) from 2009 - 2021 was utilized. The unemployment rate data of 2021 was obtained from the U.S. Bureau of Labor Statistics (Link: [Unemployment Data](https://data.bls.gov/timeseries/LNS14000000)), and the 2021 US unemployment rate was calculated by taking the average unemployment rate of each month in 2021.

This report primarily aims to address two research questions: 1) Which college degree should you pursue for a career with a high starting and mid-career salary 2) Which college degree shows the most promising growth rate from starting to mid-career salary as well as in respect to time?

## a. Set up

```{r, echo=FALSE}
# Load tidyverse package
library(tidyverse)
```

## b. Dataset Creation

```{r}
# create data frame
grad_students_data <- read.csv("~/Desktop/SDS322E/Project/project data/grad-students.csv")
recent_grads_data <- read.csv("~/Desktop/SDS322E/Project/project data/recent-grads.csv")

# remove irrelevant variables
grads <- grad_students_data %>%
  select(-"Grad_full_time_year_round", -"Nongrad_total", 
         - "Nongrad_employed", - "Nongrad_full_time_year_round",
         - "Nongrad_unemployed", - "Nongrad_unemployment_rate",
         - "Nongrad_median", - "Nongrad_P25",
         - "Nongrad_P75", - "Grad_share", 
         - "Grad_premium", - "index", - "Major_code") %>%
  arrange(desc(Grad_median)) %>%
  mutate(Rank = seq(1:173),
# add categorical variable
         Status = "Grad") %>%
  relocate(Rank, .before = Major)

# remove irrelevant variables
recent_grads <- recent_grads_data %>%
  select(- Men, - Women, - ShareWomen, - Employed, 
         - Full_time, - Part_time, - College_jobs, 
         - Non_college_jobs, -Low_wage_jobs, -index, - Rank, 
         - Major_code) %>%
  # add categorical variable
  mutate(Status = "Recent_Grad")

# rename variables
n_recent_grads <- recent_grads %>%
  rename(Employed = Full_time_year_round)

# rename variables
n_grads <- grads %>%
  select(-"Rank") %>%
  rename(Total = Grad_total, Sample_size = Grad_sample_size, Employed = Grad_employed, 
         Unemployed = Grad_unemployed, Unemployment_rate = Grad_unemployment_rate, 
         Median = Grad_median, P25th = Grad_P25, P75th = Grad_P75)

# merge datasets using major as the key
combined_data <- left_join(grads, recent_grads, by = "Major")

# reformat merged data dataset to make it tidy
new_combined_data <- rbind(n_grads, n_recent_grads)
```

# 2. Data Analysis

## a. Overall Data

### I) Average Starting and Mid-Career Median Salary

```{r}
# Distribution of average median salary of each major category by age status
new_combined_data %>%
  ggplot(aes(x = reorder(Major_category, -Median), y = Median, fill = Status)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge", width = 0.7) +
  scale_y_continuous(breaks = seq(20000, 100000, 10000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(labels = c("Graduate", "Recent Graduate")) +
  labs(title = "Average Median Salary Based on Major Category and Career Status", 
       x = "Major Category", 
       y = "Median ($)")
```

Figure 1: This plot depicts the average median salary of each major category by education status. The plot shows that on average, engineering majors have the highest starting salary as well as mid-career salary. In the next section, average median salary of starting salary and mid-career salary will be looked at independently to get a closer look at the salary distribution.

### II) Distribution of Majors in Each Major Category

```{r}
# Plot of average median starting-career salary of each major category 
n_grads %>%
  ggplot(aes(Major_category)) +
  geom_histogram(stat = "count", fill = "purple") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  scale_y_continuous(breaks = seq(0, 30, 5)) +
  labs(title = "Number of Majors per Major Category",
       x = "Major Category",
       y = "Number of Majors")

n_grads %>%
  group_by(Major_category) %>%
  summarize(count = n())
```

Figure 2: Histogram plot of the distribution of number of majors per major category. This plot shows that engineering has the most number of majors (29) while interdisciplinary has the least (1).

### III) Distribution of 'Recent-Grads' and 'Grad-students'

```{r}
# Plot of average median starting-career salary of each major category 
new_combined_data %>%
  ggplot(aes(x = Status, fill = Status)) +
  geom_histogram(stat = "count", position = "dodge") +
  theme(axis.text.x = element_text(hjust=0.5)) +
  scale_fill_discrete(labels = c("Graduate", "Recent Graduate")) +
  labs(title = "Numbers of Majors for \'Recent-Grad\' Data and \'Grad-Students\' data", 
       y = "Number of Majors",
       x = "Status")
  
new_combined_data %>%
  summarize(count_recent_grads = sum(Status == "Recent_Grad"),
             count_grads = sum(Status == "Grad"))
```

Figure 3: This plot shows the number of observations (majors) for both 'recent-grad' data and 'grad-students' data. This depicts that there were the same number of majors from each dataset: 173 majors per status.

## b. Salary Distributions

### I) Starting Salary Distribution

```{r}
# Plot of average median starting-career salary of each major category 
new_combined_data %>%
  filter(Status == "Recent_Grad") %>%
  group_by(Major_category) %>%
  mutate(avg_median = mean(Median)) %>%
  ggplot(aes(x = reorder(Major_category, -avg_median), y = avg_median)) +
  geom_bar(stat = "summary", fun = "mean", fill = "#00BFC4") +
  scale_y_continuous(breaks = seq(0, 100000, 5000), limits = c(0, 60000)) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Average Median Starting Salary of Each Major Category", 
       y = "Average Median Salary ($)",
       x = "Major Category") 

# salary summary statistics
new_combined_data %>%
  filter(Status == "Recent_Grad") %>%
  group_by(Major_category) %>%
  summarize(avg_median = mean(Median),
            P25 = (avg_median * 2) * 0.25,
            P75 = (avg_median * 2) * 0.75) %>%
  arrange(desc(avg_median))
```

Figure 4: This plot depicts the average median starting salary of each major category. Engineering majors, on average, have a significantly higher starting median salary relative to other major categories. In fact, engineering majors have a starting median salary more than \$13,000 above the next highest major category, which is business.

### II) Mid-Career Salary Distribution

```{r}
# remove use of scientific notation
options(scipen = 999)

# Plot of average median mid-career salary of each major category 
new_combined_data %>%
  filter(Status == "Grad") %>%
  group_by(Major_category) %>%
  mutate(avg_median = mean(Median)) %>%
  ggplot(aes(x = reorder(Major_category, -avg_median), y = avg_median)) +
  geom_bar(stat = "summary", fun = "mean", fill = "#F8766D") +
  scale_y_continuous(breaks = seq(0, 100000, 10000), limits = c(0, 100000))+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Average Median Mid-Career Salary of Each Major Category", 
       y = "Average Median Salary ($)",
       x = "Major Category") 

# average median salary of each major category
new_combined_data %>%
  filter(Status == "Grad") %>%
  group_by(Major_category) %>%
  summarize(avg_median = mean(Median)) %>%
  arrange(desc(avg_median))
```

Figure 5: This plot depicts the average median mid-career salary for each major category. While engineering majors are still ranked highest, physical sciences majors are ranked 2 on this plot instead of business majors. In addition, the average median salaries between consecutive major categories are much closer. Whereas the average median salary difference between the rank 1 and rank 2 major categories was above \$13,000 in the previous plot, it has reduced to less than \$6,000 in this plot (between engineering and physical sciences).

### III) Top 5 Majors of Each Major Category for Starting Salary

```{r}
# Sample code for plot of top 5 majors by median starting salary in a major category
new_combined_data %>%
  filter(Status == "Recent_Grad", Major_category == "Agriculture & Natural Resources") %>%
  arrange(desc(Median)) %>%
  slice(1:5) %>%
  ggplot(aes(x = reorder(Major, -Median), y = Median)) +
  geom_bar(stat = "summary", fun = "mean", fill = "#00BFC4") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 8)) +
  scale_y_continuous(breaks = seq(0, 50000, 10000)) +
  labs(title = "Agriculture & Natural Resources",
       x = "Major",
       y = "Median Salary ($)")

# code repeated for all major categories, only changing the major category name, and 
# x-axis scale as needed
# plots uploaded to a doc and linked in supporting text below
```

Figure 6: After visualizing the distribution of average median starting salary for each major category, majors with highest median starting salary for each of the major categories was visualized. Often times, choosing a degree and job isn't solely about the money. It is often a combination of what pays well in the field an individual is interested in. In order to answer this question, these plots depict the top 5 majors by median starting salary for each of the major categories.

*The code provided above is a sample code for just one of the major categories. Keeping the report length and flow in best interest, the plots for the other major categories have been aggregated into a google document (Link: [Starting Salary Top 5 Majors](https://docs.google.com/document/d/1Mp98-5tJW3bmnqKfa2Qe9ehO4QN_If-TYJ_UJlC34Q4/edit)) and can be viewed as intended.*

### IV) Top 5 Majors of Each Major Category for Mid-Career Salary

```{r}
# Sample code for plot of top 5 majors by median starting salary in a major category
new_combined_data %>%
  filter(Status == "Grad", Major_category == "Agriculture & Natural Resources") %>%
  arrange(desc(Median)) %>%
  slice(1:5) %>%
  ggplot(aes(x = reorder(Major, -Median), y = Median)) +
  geom_bar(stat = "summary", fun = "mean", fill = "#F8766D") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 8)) +
  scale_y_continuous(breaks = seq(0, 80000, 20000)) +
  labs(title = "Agriculture & Natural Resources",
       x = "Major",
       y = "Median Salary ($)")

# code repeated for all major categories, only changing the major category name, and 
# x-axis scale as needed
# plots uploaded to a doc and linked in supporting text below
```

Figure 7: The plots for the top 5 majors in each major category for mid-career median salary was then visualized here.

*As done previously, the code provided is a sample code for just one of the major categories. Again, keeping the report length and flow in best interest, the plots for the other major categories have been aggregated into a separate google document, (Link: [Mid-Career Salary Top 5 Majors](https://docs.google.com/document/d/1Mp98-5tJW3bmnqKfa2Qe9ehO4QN_If-TYJ_UJlC34Q4/edit)) and can be viewed as intended.*

## c. Comparison of Salary Growth

### I) Salary Growth From Starting to Mid-Career for Each Major Category

```{r, fig.width=10}
# add percent change variable: change from starting to mid-career salary
combined_data <- combined_data %>%
  group_by(Major) %>%
  mutate(percent_change = ((Grad_median - Median) / Median) * 100) %>%
  mutate(across(percent_change, round, 2))

# add percent change variable by grouping major category
dfMC <- combined_data %>%
  group_by(Major_category.x) %>%
  mutate(mean_percent_change = mean(percent_change)) %>%
  mutate(across(mean_percent_change, round, 2))

# plot of percent change of each major category
ggplot(dfMC, aes(x = reorder(Major_category.x, mean_percent_change), 
                 y = mean_percent_change)) +
  geom_bar(stat = "summary", fun = "mean", fill = "purple", alpha = 0.7) +
  geom_text(aes(label = mean_percent_change), size = 4, hjust = 1.1) +
  coord_flip() +
  labs(title = "Average Percent Growth", 
       x = "Major Category",
       y = "Percent Growth (%)")
```

Figure 8: This first plot depicts the average percent growth of median salary from starting to mid-career for each of the major categories. This plot shows that biology and life science majors have, on average, the most percent growth from starting to mid-career salary at 131.81%. Though engineering majors had the highest average median starting and mid-career salaries, this plot shows that salary growth is low, at only 68.7%.

### II) Top 20 Majors Ranked by Salary Growth

```{r, fig.height = 10, fig.width=12}
# arrange all majors by salary percent change and show top 20
dfOverall <- combined_data %>%
  arrange(desc(percent_change)) %>%
  head(20)

# plot of percent growth for the top 20 majors
options(scipen = 999)
ggplot(dfOverall, aes(x = reorder(Major, percent_change), Grad_median)) +
  geom_col(alpha = 0.5, fill = "red") +
  geom_col(aes(x = reorder(Major, percent_change), Median), alpha = 0.4, fill = "blue") +
  scale_y_continuous(breaks = seq(0, 140000, 20000)) +
  geom_text(aes(label = percent_change), size = 5, hjust = 1.1) +
  theme(axis.text.x = element_text(angle=0, size = 13), 
        plot.title = element_text(size = 13),
        strip.text = element_blank(),
        axis.text.y.left = element_text(size = 11),
        axis.title.y = element_text(size = 18),
        axis.title.x = element_text(size = 18)) +
  coord_flip() +
  labs(title = "Top 20 Majors by Percent Growth from Starting to Mid-Career Salary", 
       x = "Major",
       y = "Salary ($)")

```

Figure 9: This second plot depicts the rankings of the top 20 majors by percent growth from starting to mid-career salary. The purple colored bars represent the starting salary for the given major, and the salmon colored bars represent the mid-career salary for the major. This plot shows that of all the majors in the dataset, zoology has the highest salary growth at 323.08%. This plot supports the findings of the previous plot, as most of the majors in the top 20 rankings are in the biology and life science major category, which had the highest average percent growth in plot 8.

### III) Salary Growth of Each Major Category During 2009 - 2021

```{r, fig.height = 10, fig.width = 12}
# change Major names into lowercase to match other datasets
# create a dataset with major categories and majors only
major_cat <- n_recent_grads %>%
  mutate(major = tolower(Major)) %>%
  mutate("Major" = str_to_title(major)) %>%
  select(Major_category, Major) 

# joined the dataset of median salaries of majors from the year 2009 to 2021 with the dataset with major categories
salary_all <- read.csv("~/Desktop/SDS322E/Project/project data/Salary 2009-2021.csv")
n_salary_all <- left_join(major_cat, salary_all, by = "Major")

# used pivot_longer to condense the columns into one column labeled year
t_salary_all <- pivot_longer(n_salary_all, cols = c('X2009', 'X2010', 'X2011', 'X2012', 
                                                    'X2013', 'X2014', 'X2015', 'X2016', 
                                                    'X2017', 'X2018', 'X2019', 'X2021'), 
                             names_to = "year", 
                             values_to = "Median Salary")
# remove coerced "X"s infront of each year
t_salary_all$year <- gsub("X", "", t_salary_all$year)


# Plot of average median salary over the years 2009 - 2021 for each major category
t_salary_all %>%
  group_by(Major_category, year) %>%
  na.omit() %>%
  mutate(avg_salary = mean(`Median Salary`)) %>%
  ggplot(aes(x = as.numeric(year), y = `avg_salary`, color = Major_category)) +
  geom_point(stat = "summary", fun = "mean", size = 2) +
  geom_line() +
  scale_x_continuous(breaks = seq(2009, 2021, 1), limits = c(2009, 2021)) +
  scale_y_continuous(breaks = seq(30000, 110000, 5000)) +
  theme(axis.text.x = element_text(size = 11),
        plot.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.text.y.left = element_text(size = 11),
        axis.title.y = element_text(size = 13),
        axis.title.x = element_text(size = 13)) +
  labs(title = "Average Median Salary Over 2009 - 2021 for Each Major Category",
       x = "Year",
       y = "Average Median Salary ($)")
```

Figure 10: This plot depicts how the salary for each major category has changed from year to year over the duration of 2009 - 2021. The notable change observed in this plot is the spike in average median salary from 2012 - 2013. This time period lines up with recovery from the US housing market crash that reached a low in 2011. Overall, the percent change in salary seems to be similar over time for each of the major categories. Therefore, this data suggests that economic conditions is one of the most influential factors in median salary changes.

## d. Unemployment

### I) Unemployment Rate of Each Major Category

```{r}
# merge the datasets by Major
unem_rates_all <- read.csv("~/Desktop/SDS322E/Project/project data/Unemployment Rates 2009-2021.csv")
n_unem_rates <- left_join(major_cat, unem_rates_all, by = "Major" )

# condense the columns into one column labeled year
t_unem_rates <- pivot_longer(n_unem_rates, cols = c('X2009', 'X2010', 'X2011', 'X2012', 
                                                    'X2013', 'X2014', 'X2015', 'X2016', 
                                                    'X2017', 'X2018', 'X2019', 'X2021'), 
                                           names_to = "year", 
                                           values_to = "Unemployment Rates")
# remove coerced "X"s infront of each year
t_unem_rates$year <- gsub("X", "", t_unem_rates$year)

# filter for year 2021 and save dataset
c_unem_rates <- t_unem_rates %>%
  filter(year == 2021) %>%
  na.omit() %>%
  group_by(Major_category) %>%
  mutate(avg_unem_rate = mean(`Unemployment Rates`)) %>%
  mutate(across(avg_unem_rate, round, 2))

# download US unemployment data
USunem <- read.csv("~/Desktop/SDS322E/Project/project data/US_unemployment.csv")
# pivot table to make it tidy and rename
USunem <- USunem %>%
  pivot_longer(cols = c("Jan":"Dec")) %>%
  rename(Month = "name", 
         Unemployment_rate = "value")
# calculate average US unemployment rate 
USunem %>%
  filter(Year == 2021) %>%
  summarize(avg = mean(Unemployment_rate))
  
# plot of average unemployment rate in 2021 for each major category with verticle line to    # represent overall US unemployment rate in 2021
c_unem_rates %>%
  ggplot(aes(x = reorder(Major_category, avg_unem_rate), 
             y = avg_unem_rate)) +
  geom_bar(stat = "summary", fun = "mean", fill = "blue", alpha = 0.5) +
  geom_hline(yintercept=5.37,color="red") +
  geom_text(aes(label = avg_unem_rate), hjust = 1.1) + 
  scale_y_continuous(breaks = seq(0, 8, 0.5)) +
  coord_flip() + 
  labs(title = "Average Unemployement Rate of Each Major Category",
       x = "Major Category",
       y = "Average Unemployment Rate (%)")

# summary statistics
# categorize average unemployment as high if above national average, low if below
c_unem_rates %>%
  group_by(Major_category) %>%
  summarize(avg_unemployment_rate = mean(`Unemployment Rates`),
            ifelse(avg_unemployment_rate > 5.37, "High", "Low")) %>%
  arrange(desc(avg_unemployment_rate)) %>%
  rename(Unemployment_Severity = "ifelse(avg_unemployment_rate > 5.37, \"High\", \"Low\")") 
```

Figure 11: This plot depicts the average unemployment of each major category in 2021. The red vertical line represents the overall unemployment rate in the US in 2021 (5.37%). This plot indicated arts majors were the only major category to have an unemployment rate above the national average. This is likely because careers in art may not have typical employers. On the other hand, all the other major categories had unemployment rates below the national average. This is likely due to the fact that having some sort of degree places an individual at an advantage when obtaining a job compared to an individual without a college degree.

# 3. Discussion

## a. Research Question 1: Which college degree should you pursue for a career with a high starting and mid-career salary?

The data indicated that pursuing a degree in an engineering major is most beneficial to obtain a high starting as well as mid-career salary, as depicted in figures 4 and 5. The average median starting salary for the engineering major category was \$57, 382, and the average median mid-career salary was \$94,327.59 (2010 - 2012). However, for individuals who have interests in other fields, it would be more helpful to look at the top paying majors for each major category. For example, as depicted in figure 6, those who have an interest in the agriculture & natural resources field should consider a degree in food science to prioritize a high starting salary. On the other hand, as depicted in figure 7, those prioritizing a high mid-career salary (in the same field) should consider a degree in agricultural economics.

## b. Research Question 2: Which college degree shows the most promising growth rate from starting to mid-career salary as well as in respect to time?

A degree in biology and life sciences was shown to be have the highest growth rate from starting to mid-career median salary at 131.81% (figure 8). When compared with all majors in the dataset, zoology was shown to have the highest percent growth at 323.08% (figure 9), which unsurprisingly is in the biology and life sciences major category.

In terms of salary growth in respect to time, the data showed that salary growth was similar for the major categories during the years 2009 - 2021 (figure 10). Much of the pattern of salary change for each major category as shown in figure 10 is similar, including the notable spike in median salary that all major categories experienced in 2012. Therefore, as noted prior, this suggests that overall economic trends are a strong predictor for median salary growth from year to year for most majors.

## c. Reflections

The most valuable skill I learned from this project is the process of defining a research question(s) to investigate, finding the data necessary to address the problem, and designing the work flow to execute the report. This process was time consuming and difficult at times, but finding ways to overcome these challenges gave me valuable skill sets I can apply in the future. The most challenging aspect of this project was defining the research question as well as finding the data itself. In addition, methods of coding we have not covered in class often had to be utilized in this project, which was challenging and stressful at times, but discovering ways to overcome them was one of the reasons that made this project fun.
